### 2025-08-27 sections/3-datasets.tex
added lfd discussion

>>>>

Since the time complexity of CAKES algorithms scales with the LFD of the dataset, we examine the LFD of each
dataset we used for benchmarks, as shown in Figure~\ref{fig:results:lfd-plots}. In this section, when we discuss trends in LFD, unless otherwise noted, we are referring to
the 95th percentile of LFD.

In Figure~\ref{fig:results:fashion-mnist-lfd}, we observe that until approximately depth 5, Fashion-MNIST’s LFD is low (i.e., less than 4). 
It then starts increasing, reaching a peak of about 6 near depth 20, before decreasing to 1 at the maximum depth.
Relative to Fashion-MNIST, Glove-25 has low LFD, as shown in Figure~\ref{fig:results:glove-25-lfd}. 
In particular, Glove-25’s LFD is less than 3 for all depths.
Figure~\ref{fig:results:sift-lfd} shows the LFD by depth for Sift, which has higher LFD relative to Fashion-MNIST and Glove-25. 
It increases sharply to a peak of 9 around a depth of 10.
For the Random dataset, which has the same cardinality and dimensionality as Sift, LFD starts at 20 at depth
0 and all percentile lines decrease linearly with depth until reaching the leaves of the tree. The spread in LFD starts very
small for the first few clusters and increases as depth increases. For Silva, as
shown in Figure~\ref{fig:results:silva-lfd}, exhibits consistently low LFD. In particular, LFD is less than 3 for all depths, hovering near 1 for
clusters at depth 40 and deeper. For the Radio-ML dataset, Figure~\ref{fig:results:radioml-lfd} LFD values show three distinct peaks around an LFD
of 12 at or near depths of 8, 25 and 50. Each peak is followed by a linear decrease until encountering a sharp spike for
the next peak. Within each of the three portions, this dataset has a character very similar to that of the Random dataset.

### 2025-08-27 sections/5-discussion.tex
added lfd discussion

We have presented CAKES, a suite of three algorithms for fast $k$-NN search across diverse distance functions. When the distance is a metric (Sec.~\ref{sec:methods}), the algorithms are exact; under cosine distance (non-metric) they achieve nearly perfect recall. CAKES is most effective when data satisfy the manifold hypothesis—lying on a low-dimensional manifold within a high-dimensional space. It builds an unbalanced-clustering binary tree that captures manifold structure and accelerates search. Conversely, performance degrades on randomly distributed data, where such structure is absent.

@@

We stress that CAKES targets \emph{big} data: linear search is better at small sizes, but for each ANN-Benchmarks dataset tested, CAKES overtakes linear search at sufficiently large cardinality—about $10^{5}$ for Fashion-MNIST and $10^{6}$ for Glove-25. On the high-LFD Random dataset, CAKES never beats linear search. These results support that CAKES performs well on data arising from constrained generating phenomena even as cardinality grows exponentially.

>>>>

We have presented CAKES, a suite of three algorithms for fast $k$-NN search across diverse distance functions. When the distance is a metric (Sec.~\ref{sec:methods}), the algorithms are exact; under cosine distance (non-metric) they achieve nearly perfect recall. CAKES is most effective when data satisfy the manifold hypothesis—lying on a low-dimensional manifold within a high-dimensional space. It builds an unbalanced-clustering binary tree that captures manifold structure and accelerates search. Conversely, performance degrades on randomly distributed data, where such structure is absent. In Figure~\ref{fig:results:lfd-plots}, we show the extent to which each dataset we texted exhibits a manifold structure, as quantified by the percentiles of LFD.

@@

We stress that CAKES targets \emph{big} data, and our results support this claim; we
observe that while linear search can outperform CAKES on datasets with small cardinality, CAKES always overtakes
linear search at a large enough cardinality for each of the ANN-benchmark datasets we tested. 
Moreover, we observe
that CAKES’s algorithms outperform linear search at a lower cardinality when the dataset has a lower LFD. For example,
Sift has much higher LFD than Fashion-Mnist and Glove-25, and we observe that CAKES overtakes linear search at
a cardinality of $10^{7}$ 
for Sift, as opposed to about $10^{5}$
and $10^{6}$ 
for Fashion-Mnist and Glove-25, respectively. For the
Random dataset, which has the highest LFD, CAKES’s algorithms \emph{never} outperform linear search.
